\Large \bf 6.824 Final Project

### 2014-05-12 writeup/paper.tex
Merge with new server config section

\subsection{Persistent Storage}
In order to tolerate servers restarting and to handle databases that
cannot fit in memory, all three components of the system store state
and data persistently on disk.  To interface with the disk, a Go
wrapper for the high-performance C++ key/value libray LevelDB was
utilized.  An outline of what each system component stores on disk is
presented below.
\begin{itemize}
\item The Paxos processes write the highest accepted value, the
  highest accepted proposal number, and the highest acknowledged
  prepare request to disk for each instance.  This ensures that the
  protocol will operate correctly even if a server restarts (for
  example, a decided value can never be changed).  In addition to this
  state, the Paxos processes persistently store the done responses
  received from peers and the number of the maximum known instance
  (see the recovery section below).

@@

\begin{itemize}
\item The Paxos process will load any saved done responses and maximum
  instance number from its disk.  It will then contact peers to
  request their done responses and maximum instance number,
  determining which peer is the most updated out of any peers that
  respond.  From these values, both the ones read from disk and the
  ones received, it can determine which instances it should know about
  but does not; it then requests these instances from its peers.
\item The shard master process will load the highest known
  configuration and highest executed Paxos log instance from disk if
  it exists.  It will then request the same information from its
  peers, determining who among the responders is the most updated.
  From this information it can determine if there are any missed
  configurations, and if so request them from a peer.
\item The shard key/value service will load its last known
  configuration and the highest executed Paxos log instance from disk
  if it exists.  It will then request the same information from peers,
  determining the most updated state that it should have.  Finally, it
  can request the actual database key/value entries from a peer (see
  below for details of how this transfer is performed) along with any
  client responses and lists of seen operations.
\end{itemize}
These protocols allow any number of servers restarting with their disk
contents to catch themselves up from any updated peer.  They also
allow any number of servers restarting without their disk contents to
recover the needed data from any peer that has the data.

\textbf{Single server crashing and restarting, disk in-tact or lost}:
The server will contact its peers to figure out the state it is
missing, and it will then send RPCs to collect the missing state. The
server it is doing state transfer with cannot change disk state until
the entire process is done. If the disk is lost, the request indicates
that it has no state, which causes the entire database to be sent.

\textbf{Complete crash of all servers}: Servers read their state from
disk, and then contact peers to figure out if they are out of date.

%% \subsubsection{Memory and Performance Considerations}
%% There are a number of design choices which were made in order to
%% address both memory management and overall performance.

%% Firstly, the service must be able to handle database content which
%% is too large to fit in volatile memory.  Towards this end, values
%% are written to disk immediately as described above so it does not
%% need to be kept in memory.  More importantly, however, servers that
%% need to transfer shards (either because the shard master decided a
%% shard needs to move or because a server is recovering) must be able
%% to do so in multiple RPCs since memory is too small to create a
%% reply with the entire database.  Therefore, a shard transfer
%% occurrs with the following protocol:
%% \begin{itemize}
%% \item The requester sends a request to a peer, including a request
%%   number, and the current configuration number.  If doing a shard
%%   transfer, the shard number is also included.
%% \item If the receiver is currently at the appropriate
%%   configuration, it accepts the request.  It will then begin
%%   populating a reply with client response data, a list of seen
%%   operations, and the desired key/value data.  While doing this, it
%%   will periodically check memory usage; if it exceeds a
%%   configurable threshold, it will reply with whatever it currently
%%   has and mark the reply as incomplete.
%% \item The requester copies the data from the reply to its disk.  If
%%   the reply was marked as incomplete, it sends another request to
%%   the same peer with an incremented request number.
%% \item This continues until the reply is marked as complete, at
%%   which point the requester launches a goroutine to make sure the
%%   sender receives an acknowledgement of the receipt of the final
%%   data message.
%% \item If at any point the requester decides that the contacted peer
%%   is dead, it launches an acknowledgement goroutine as previously
%%   described for that peer (to terminate the transfer in case the
%%   peer wakes up or is not dead) and moves on to try a different
%%   peer, resetting the request number to 0.
%% \end{itemize}
%% Both the requester and the sender ignore client requests until the
%% transfer is complete - this ensures that the requester will receive
%% a consistent snapshot of the data (and is the reason the requester
%% launches acknowledgement goroutines since otherwise a peer
%% erroneously declared dead would be halted forever).

One aspect which addresses performance is that progress can still
be made while recovery is taking place as long as a majority of
servers are not recovering.  For example, if a replica group has
three servers and a shard transfer is ocurring between two of them
(perhaps because one just restarted), a client can still have its
requests processed if it contacts the third peer.  This is because
the Paxos service runs independently of the shard key/value
service; although the latter will ignore client requests on the
transferring servers, their Paxos processes can still respond and
make progress.  Since the Paxos instances are stored persistently
on disk, the two occupied servers can catch themselves up from the
log when they complete the transfer.

@@

When one server contacts another via the FetchRecovery RPC, both
servers enter a recovering state. This state means that neither are
allowed to modify their databases \emph{except} through the
recovery process. Depending on how many replicas there are, this
may render a replica group unable to process requests while
recovery is underway.  If there are enough replicas so that paxos
majority can still be reached, then the recovering servers will
catch up once they come back online, using the functionality
already built in to sync lagging paxos servers.

All recovery processes send over both the relevant portions of the
k/v store, as well as paxos state and any state needed to detect
duplicates.

%% We use leveldb. To guarantee persistence, we enforce writing to
%% disk all operations before returning. This slows performance down,
%% but prevents loss of data that the client wants committed. Speed
%% could be gained using a Harp-like set of pointers, but this would
%% be complicated and the client would have to understand the modified
%% persistence model.

@@

\subsection{Server Configuration}

Two Amazon AWS clusters were provisioned to test real-life performance of the system. The faster of the clusters used Amazon's m3.large instance type, where each server has 7.5 gigabytes of RAM, 2 Intel Xeon i5 Sandy Bridge CPU cores, and 1 32 gigabyte solid state drive. Real-world network performance, measured using iperf, was approximately 700 Mb/s. The slower of the clusters used Amazon's m1.small instance type, where each server has 1.7 gigabytes of RAM, 1 Intel Xeon Westmere core, and 1 160 gigabyte hard disk drive. Real-world network performance was approximately 150 Mb/s. Each cluster had one control node and twelve slave nodes. Instances were allocated on multi-tenant hardware, and nodes within a cluster were within the same region and availability zone.

@@

\subsection{Bottlenecks}
Although the above enhancements significantly improved performance,
various were identified which could be addressed in the future.

One notable bottleneck is disk reading and writing.  The database
library wrapper, LeveDB, offers a cache as well as compression and is
quite fast.  However, as discussed previously, benchmark tests
revealed that scanning the database to compile data for shard
transfers accounted for a significant portion of total transfer time.
Additionally, in the process of debugging networked RPCs, simple
logging was implemented which slowed down operations by nearly a
factor of ten (it was disabled for benchmarks); this was a good
demonstration of disk latency and how care must be taken in even
mundane aspects of the system.  In the future, the system topology
could be somewhat altered to allow for more parallelizable recovery
and transfer procedures to help minimize the effect of disk latency
(for example, a topology inspired by FDS).

For disk recovery, we also examined the possibility of using a raw TCP
stream to send the data, instead of using RPCs, in order to reduce
overhead. The transfer rate for a 1GB database was about 4.5
Mbps. This was approximately equivalent to the data rate at which the
database itself was created locally. Thus, the overhead necessary to
read and write values to the database is a key bottleneck.

Another bottleneck is network capacity and/or number of messages.  As
shown in the latency and thoughput section, increasing the number of
replicas has an adverse effect on performance since each operation
requires communication with a majority of replicas.  The paxos
optimizations therefore helped to improve performance, and further
optimizations could be made to the data transfer protocols to reduce
the required number of messages.

Increasing the number of replica groups increases the performance of
the system for normal operation, especially under heavy load with
evenly distributed key requests. This is especially noticable in the
system throughput metric.  This arises since number of groups roughly
corresponds to how parallelizable the operations are.

>>>>

\subsection{Persistent Storage}
In order to tolerate servers restarting and to handle databases that
cannot fit in memory, all three components of the system store state
and data persistently on disk.  To interface with the disk, a Go
wrapper for the high-performance C++ key/value libray LevelDB was
utilized.  An outline of what each system component stores on disk is
presented below.
\begin{itemize}
\item The Paxos processes write the highest accepted value, the
  highest accepted proposal number, and the highest acknowledged
  prepare request to disk for each instance.  This ensures that the
  protocol will operate correctly even if a server restarts \cite{paxos} (for
  example, a decided value can never be changed).  In addition to this
  state, the Paxos processes persistently store the done responses
  received from peers and the number of the maximum known instance
  (see the recovery section below).

@@

\begin{itemize}
\item The Paxos process will load any saved done responses and maximum
  instance number from its disk.  It will then contact peers to
  request their done responses and maximum instance number,
  determining which peer is the most updated out of any peers that
  respond.  From these values, both the ones read from disk and the
  ones received, it can determine which instances it should know about
  but does not; it then requests these instances from its peers.
\item The shard master process will load the highest known
  configuration and highest executed Paxos log instance from disk if
  it exists.  It will then request the same information from its
  peers, determining who among the responders is the most updated.
  From this information it can determine if there are any missed
  configurations, and if so request them from a peer.
\item The shard key/value service will load its last known
  configuration and the highest executed Paxos log instance from disk
  if it exists.  It will then request the same information from peers,
  determining the most updated state that it should have.  Finally, it
  can request the actual database key/value entries from a peer (see
  below for details of how this transfer is performed) along with any
  client responses and lists of seen operations.
\end{itemize}

These protocols allow any number of servers restarting with their disk contents to catch themselves up from any updated peer.  They also allow any number of servers restarting without their disk contents to recover the needed data from any peer that has the data.

\textbf{Single server crashing and restarting, disk intact or lost}: As described above, each
recovery process contacts a random replica in the same replica group.  The recovering server can then get any missing data by comparing states with the updated server (e.g. shard data, shard master configurations, and Paxos instances - note that Paxos only discards old decisions if all participants have marked that sequence number as 'done').  This will therefore work regardless of whether the disk was lost or simply out of date.  Note that when data is transferred, multiple messages may be used for a single shard if it is too large to fit in memory - the sender will periodically check memory usage while constructing the reply and mark replies as incomplete if needed.

% Is the below actually implemented?
%The recovery process contacts a random replica in the same replica group, and tells the up-to-date server that it has no state. The two servers open an auxiliary TCP connection to send over the relevant portions of the database. This avoids RPC overhead, since the database might be large. The database has an in-order iterator, in case the TCP connection fails, the recoverer can re-establish the connection and ask the sender to pick up where it left off. If the sending server fails, then the recovering server asks a new replica. This allows for a failure mode where the original server remains locked permanently. We have decided that this is preferable to sacrificing consistency, and the system administrator can resolve the problem by rebooting the hung server.

\textbf{Complete crash of all servers}: If all servers lose their disk state, then all will restart and begin like a brand new deployment (the above protocols will simply not find any updated data to transfer). If the disks are not lost, then the recovery process will read the various states written to disk and pick up where it left off.  Because disk writes take place before any operation commits, data is not lost.

It should be noted that during a shard transfer (including recovery), the shard key/value service on both the sender and recoverer ignore client requests in order to ensure that consistent data is sent (their databases cannot be modified except through the recovery process itself).  Since their Paxos processes continue running independently, however, progress can still be made even if a majority of servers is involved in a data transfer (the decided operations will be persistently logged and the halted servers can catch up when transfer is complete).  This therefore improves performance during recoveries or transfers.

@@

@@

\subsection{Server Configuration}
Two Amazon AWS clusters were provisioned to test real-life performance of the system. The faster of the clusters used Amazon's m3.large instance type, where each server has 7.5 gigabytes of RAM, 2 Intel Xeon i5 Sandy Bridge CPU cores, and 1 32 gigabyte solid state drive. Real-world network performance, measured using iperf, was approximately 700 Mb/s. The slower of the clusters used Amazon's m1.small instance type, where each server has 1.7 gigabytes of RAM, 1 Intel Xeon Westmere core, and 1 160 gigabyte hard disk drive. Real-world network performance was approximately 150 Mb/s. Each cluster had one control node and twelve slave nodes. Instances were allocated on multi-tenant hardware, and nodes within a cluster were within the same region and availability zone.

@@

\subsection{Bottlenecks}

Although the above enhancements significantly improved performance, various were identified which could be addressed in the future.

One notable bottleneck is disk reading and writing.  The database library wrapper, LeveDB, offers a cache as well as compression and is quite fast.  However, as discussed previously, benchmark tests revealed that scanning the database to compile data for shard transfers accounted for a significant portion of total transfer time.  Additionally, in the process of debugging networked RPCs, simple logging was implemented which slowed down operations by nearly a factor of ten (this was disabled for the above benchmarks); this was a good demonstration of disk latency and how care must be taken in even mundane aspects of the system.  It was considered adding a commit point marker (similar to the ideas presented in \cite{harp}), which would be straightforward in the current system by simply returning before the disk write and not advancing the Paxos min until the write completes, but the above measurements indicate that database scanning is much more significant than isolated accesses.  In the future, the improvements can be made to avoid this iteration bottleneck.  For example, the system topology could be altered to allow for more parallelizable recovery with a procedure inspired by FDS \cite{fds}.  More simply, however, the servers can be altered to store a separate database for each shard; the entire database could then be transferred to the receiver using TCP/UDP.  Preliminary tests on AWS showed that a 100MB database can be sent over TCP in less than one second.  In addition, this would avoid the need for either the sender or the receiver to scan through a database, thus eliminating the major bottleneck identified previously.

Another bottleneck is network capacity and/or number of messages.  As shown in the latency and throughput section, increasing the number of replicas has an adverse effect on performance since each operation requires communication with a majority of replicas.  The paxos optimizations therefore helped to improve performance, and further optimizations could be made to the data transfer protocols to reduce the required number of messages.

Increasing the number of replica groups increases the performance of the system for normal operation, especially under heavy load with evenly distributed key requests. This is especially noticeable in the system throughput metric.  This arises since number of groups roughly corresponds to how parallelizable the operations are.
